---
title: "A brief introduction to BamLiner"
author:
- name: "Roberto Amarie"
  affiliation: Università degli studi di Milano & Politecnico di Milano, Italy
  email: roberto.amarie@mail.polimi.it    or    roberto.amarie@studenti.unimi.it  
date: "`r Sys.Date()`"    
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{BamLiner}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

@unsure if including these first 2 chunks

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, 
  comment = "#>")
```

```{r setup}
library(BamLiner)
```

------------------------------------------------------------------------

```{r style, echo = FALSE, results = 'asis'}
library(BiocStyle)
```

```{r, echo = FALSE}
library(knitr)
```

*Notes*:

# Introduction

blablablablabla blablablablabla blablablablabla blablablablabla
blablablablabla blablablablabla blablablablabla blablablablabla
blablablablabla

------------------------------------------------------------------------

# Installing and loading the package {#installation}

## Installation

### Bioconductor

**BamLiner** requires several CRAN and Bioconductor R packages to be
installed. Dependencies are usually handled automatically, when
installing the package using the following commands:

```         
install.packages("BiocManager")
BiocManager::install("BamLiner")
```

[NOTE: Ignore the first line if you already have installed the
`r CRANpkg("BiocManager")`.]

### Manual installation

In the unlikely case that a manual installation is required, e.g., if
you do not install **BamLiner** via Bioconductor (which is highly
recommended), before installing **BamLiner** make sure the following
packages are installed:

<!-- CRAN packages: -->

<!-- `r CRANpkg("Matrix")`, `r CRANpkg("quadprog")`, `r CRANpkg("plyr")`, -->

<!-- `r CRANpkg("ggplot2")`, `r CRANpkg("ggseqlogo")`, -->

<!-- `r CRANpkg("gridExtra")` -->

Bioconductor packages:

`r Biocpkg("GenomicRanges")`, `r Biocpkg("Biostrings")`,
`r Biocpkg("BiocGenerics")`, `r Biocpkg("S4Vectors")`,
`r Biocpkg("BSgenome.Hsapiens.UCSC.hg38")`,
`r Biocpkg("GenomicFeatures")`, `r Biocpkg("GenomeInfoDb")`, \#### maybe
these last 2? have to check

CRAN packages can be installed from R using the following command:

```         
install.packages("<package_name>")
```

Bioconductor packages can be installed from R using the following
command:

```         
BiocManager::install("<package_name>")
```

Sometimes, it may also be useful to update Bioconductor:

```         
BiocManager::install()
```

Finally, the manual installation of **BamLiner** can, for example, be
done from the command line ...

```         
R CMD INSTALL BamLiner_<version>.tar.gz
```

... or the newest version can directly be installed from GitHub using
the CRAN package `r CRANpkg("devtools")`:

```         
library(devtools)
install_github("rmpiro/BamLiner")       #######well maybe I have to remove this part if I don't get to push it on my github
```

## Loading the package

After installation, loading the package is simple:

```{r, results='hide', message=FALSE}
library(BamLiner)
```

------------------------------------------------------------------------

# Input data {#input}

**BamLiner** works with SAM/BAM files resulting from any genome
sequencing experiment, except for those including SA:Z tagged chimeric
reads. For detailed information on this format (including an example),
see <https://samtools.github.io/hts-specs/SAMv1.pdf>.

Specifically, it requires that the files should be pre-loaded in R in
the form of a `r Rpackage("GAlignments")` object.

Additionally, **BamLiner** requires the genomic sequence (in form of a
`r Rpackage("BSgenome")` object) to determine the reference sequences on
which reads should map to.

## Bam/Sam file {#input-file}

Here show how to import the GAlignment object and which mcols() to
include

**Example workflow:**

```{r}
# load example signatures for breast cancer genomes from Nik-Zainal et al
# (PMID: 22608084) in the format produced by pmsignature (PMID: 26630308)
pmsigdata <- system.file("extdata",
          "Nik-Zainal_PMID_22608084-pmsignature-Param.Rdata", 
          package="decompTumor2Sig")
load(pmsigdata)
 
# extract the signatures from the pmsignature 'Param' object
signatures <- getSignaturesFromEstParam(Param)
signatures[1]
```

**Example workflow:**

```{r}
# load the reference genome and the transcript annotation database
refGenome <- BSgenome.Hsapiens.UCSC.hg19::BSgenome.Hsapiens.UCSC.hg19
transcriptAnno <-
           TxDb.Hsapiens.UCSC.hg19.knownGene::TxDb.Hsapiens.UCSC.hg19.knownGene

# take the six breast cancer genomes from Nik-Zainal et al (PMID: 22608084) 
gfile <- system.file("extdata",
                     "Nik-Zainal_PMID_22608084-VCF-convertedfromMPF.vcf.gz",
                     package="decompTumor2Sig")

# read the cancer genomes in VCF format
genomes <- readGenomesFromVCF(gfile, numBases=5, type="Shiraishi",
                              trDir=TRUE, refGenome=refGenome,
                              transcriptAnno=transcriptAnno, verbose=FALSE)
length(genomes)
genomes[1:2]
```

**Example workflow:**

```{r}
# load the reference genome and the transcript annotation database
refGenome <- BSgenome.Hsapiens.UCSC.hg19::BSgenome.Hsapiens.UCSC.hg19
transcriptAnno <-
           TxDb.Hsapiens.UCSC.hg19.knownGene::TxDb.Hsapiens.UCSC.hg19.knownGene

# take the six breast cancer genomes from Nik-Zainal et al (PMID: 22608084) 
gfile <- system.file("extdata", "Nik-Zainal_PMID_22608084-MPF.txt.gz", 
                     package="decompTumor2Sig")

# read the cancer genomes in MPF format
genomes <- readGenomesFromMPF(gfile, numBases=5, type="Shiraishi",
                              trDir=TRUE, refGenome=refGenome,
                              transcriptAnno=transcriptAnno, verbose=FALSE)
```

### Verifying the signature format

For certain applications it may be necessary to construct signatures or
convert them from other kind of data. To do so, each signature must be
either a numeric *vector* of probabilities which sum up to 1 (for
Alexandrov-type signatures) or a *matrix* or *data.frame* with six
numeric columns, every row of which sums up to 1 (for Shiraishi-type
signatures).

A set of signatures is then simply a *list* of such signatures.

To verify whether a signature set has a format that can be used with
**decompTumor2Sig** the package provides the following set of functions.
Since the mutation frequencies in genomes are specified in exactly the
same way, these functions can be used for both signatures and genomes:

-   **isAlexandrovSet()**: Verify whether a list object (set of
    signatures or genomes) is compatible with the Alexandrov model.
-   **isShiraishiSet()**: Verify whether a list object (set of
    signatures or genomes) is compatible with the Shiraishi model.
-   **isSignatureSet()**: Verify whether a list object (set of
    signatures or genomes) is compatible with either the Alexandrov or
    the Shiraishi model.
-   **sameSignatureFormat()**: Verify whether two list objects (two sets
    of signatures or genomes) contain signatures/genomes of the same
    format.

Examples:

```{r}
isAlexandrovSet(sign_a)
isSignatureSet(sign_a)
isShiraishiSet(sign_s)
isSignatureSet(sign_s)
sameSignatureFormat(sign_a, sign_s)
```

------------------------------------------------------------------------

# Workflow {#workflow}

(*Note*: The following examples are for illustrative purpose only and
need not be biologically meaningful.)

Once both the tumor genome(s) and the given mutational signatures have
been read (see above), the contribution of the given signatures to the
somatic mutations in individual tumors can be determined using the
following workflow.

In the following, we assume to have the signatures in a list object
named *signatures* and the mutation frequencies of the tumor genome(s)
in another list object named *genomes*.

**Important note**: it is imperative that the mutation frequencies
represented by both *signatures* and *genomes* have been computed with
the same characteristics. That is, if the *signatures* refer to
sequences of 5 nucleotides/bases (with the mutated base at the center),
then also the *genomes* must have been read for 5 bases. If the
*signatures* have been produced taking the transcription direction into
account, then also the *genomes* must have been read taking the
transcription direction into account, and so on.

### Example: input data

As a simple example, load a set of 15 Shiraishi-type signatures (object
*signatures*) provided with this package. These signatures were obtained
with the package `r Rpackage("pmsignature")` from a set of 435 tumor
genomes with at least 100 somatic mutations from ten different tumor
entities (data from Alexandrov et al. Nature 500:415-421, 2013; for the
analysis, see our papers mentioned in Section 1.1):

```{r}
# load the 15 Shiraishi signatures obtained from 
# 435 tumor genomes from Alexandrov et al.
sfile <- system.file("extdata",
              "Alexandrov_PMID_23945592_435_tumors-pmsignature-15sig.Rdata", 
              package="decompTumor2Sig")
load(sfile)
length(signatures)
signatures[1]
```

This loads an object *signatures* with 15 Shiraishi signatures for
mutated subsequences of length 5 (five nucleotides with the mutated base
at the center) and taking transcription direction into account.

Now read the tumor genomes (object *genomes*) provided with this
package, as described in Section 3.2.1:

```{r}
# load the reference genome and the transcript annotation database
refGenome <- BSgenome.Hsapiens.UCSC.hg19::BSgenome.Hsapiens.UCSC.hg19
transcriptAnno <-
           TxDb.Hsapiens.UCSC.hg19.knownGene::TxDb.Hsapiens.UCSC.hg19.knownGene

# take six breast cancer genomes from Nik-Zainal et al (PMID: 22608084) 
gfile <- system.file("extdata",
                     "Nik-Zainal_PMID_22608084-VCF-convertedfromMPF.vcf.gz", 
                     package="decompTumor2Sig")

# read the cancer genomes in VCF format
genomes <- readGenomesFromVCF(gfile, numBases=5, type="Shiraishi",
                              trDir=TRUE, refGenome=refGenome,
                              transcriptAnno=transcriptAnno, verbose=FALSE)
```
